\bf The forestecology R package for fitting and assessing models of interspecies competitive effects on the growth of trees

### 2021-03-06 paper/forestecology.Rmd
first batch of edits

abstract: 
- |
  | 1. Many models for the growth of trees that incorporate interspecies competition are based on a neighborhood effect assumption whereby all trees within a fixed distance of a focal tree are considered competitors. Methods and tools are needed to quantify this competitive effect and assess the quality of all resulting models. 
  | 2. We present the `forestecology` package providing methods for both 1) evaluating the effect of competitor species identity using permutation tests and 2) evaluating model performance using spatial cross-validation.  Following @allen_permutation_2020, we implement a Bayesian linear regression competition model.
  | 3. We demonstrate the package's functionality using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO global network of reseach sites. Given ForestGEO's data collection protocols and data formatting standards, the package was designed with cross-site compatibility in mind. We demonstrate that both 1) competitor species identity matters and 2) that not spatially cross-validating leads to error estimates that are overly optimistic. 
  | 4. The package features 1) `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence, 2) functions with standardized inputs/outputs of simple features `sf` package class, and 3) an S3 object-oriented implementation of the Bayesian linear regression model. These three facts allow for clear articulation of all the steps in the sequence of analysis and easy wrangling and visualization of the geospatial forestry data. Furthermore, while the package only has Bayesian linear regression implemented, the package was designed with extensibility to other methods in mind.

keywords:
- forest ecology, competition, tree growth, R, tidyverse, sf, cross-validation, spatial statistics

@@

BERT: Dave could you write a few sentences (and add references where appropriate) starting with the big picture of modeling tree growth, zooming in to neighhorhood models, and then connecting to the "@allen_permutation_2020 considers the following model" I write below?

Repeat-censused forest plots offer excellent data to test neighborhood models of the effect of competition on the growth of trees @canham_neighborhood_2006 @uriarte_spatially_2004.

@@

Figure \ref{fig:scbi-simulation}

@@

# Discussion

The `forestecology` package provides

@@

AYK and DNA conceived the ideas and coded most of the package in the early design stages. AYK led the writing of the manuscript. SPC refactored much of the package's code in the later design stages to align with R and "tidy" programming best practices [@wickham_welcome_2019]. All authors contributed critically to the drafts and gave final approval for publication.

>>>>

abstract: 
- |
  | 1. Neighborhood competition models are powerful tools to measure the effect of interspecific competition. Statistical methods to ease the application of these models are currently lacking.
  | 2. We present the `forestecology` package providing methods to i) specify neighborhood competition models, ii) evalulate the effect of competitor species identity using permutation tests, and iii) measure model performance using spatial cross-validation.  Following @allen_permutation_2020, we implement a Bayesian linear regression neighborhood competition model.
  | 3. We demonstrate the package's functionality using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO global network of reseach sites. Given ForestGEO's data collection protocols and data formatting standards, the package was designed with cross-site compatibility in mind. We demonstrate that both i) competitor species identity matters and ii) that not spatially cross-validating leads to error estimates that are overly optimistic. 
  | 4. The package features i) `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence, ii) functions with standardized inputs/outputs of simple features `sf` package class, and iii) an S3 object-oriented implementation of the Bayesian linear regression model. These three facts allow for clear articulation of all the steps in the sequence of analysis and easy wrangling and visualization of the geospatial forestry data. Furthermore, while the package only has Bayesian linear regression implemented, the package was designed with extensibility to other methods in mind.

keywords:
- forest ecology, interspecific competition, neigbhorhood competition, tree growth, R, ForestGEO, cross-validation, spatial statistics,

@@

Repeat-censused forest plots offer excellent data to test neighborhood models of the effect of competition on the growth of trees (@canham_neighborhood_2004). Neighborhood models of competition have been used to: test whether the species identity of a competitor matters (@uriarte_spatially_2004); measure species-specific competition coefficients (@das_effect_2012 @tatsumi_estimating_2016); test competing models to see what structures competitive interactions, e.g. traits or phylogeny (@allen_permutation_2020 @uriarte_trait_2010); and inform selective logging practices (@canham_neighborhood_2006). Although these are well-described methods, no methods are currently available for their easy application. Here we address this in an R package. We largely follow the methods presented in @allen_permutation_2020. The package is written to model stem radial growth between two censuses based on neighborhood competition. It assumes a single large plot using the ForestGEO plot data structure as a guide (@andersonteixeira_ctfs-forestgeo_2015).

@@

Figure \ref{fig:scbi-simulation} highlights the importance of spatial cross validating when interpreting results from neighborhood compeition models. Without cross-validation the competitor idenity did matter, the non-permuted competitor species had a much lower RMSE than the permuted one. But once we include the spatial cross validation this improvement disaprees. So in this 9 ha subplot of the SCBI plot competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other places (@allen_permutation_2020 @uriarte_spatially_2004).

@@

# Conclusion

The `forestecology` package provides an accessible way to fit and test models of neighborhood competition. Currently it is written to work with data from ForestGEO plots. But it could easily be modified to work on any single large, mapped forest plot in which at least two measurements of each individual have been taken. With a little bit of work the package could also be applied to forest inventory data in which several small plots are spatially separated, e.g., USFS Forest Inventory. In a future version of the package we also hope to make it possible to model plant mortality in addition to plant growth. The package follows the guidelines for `tidy` data, `sf` spatial structure., and S3 open-oriented model structure. These make for ease of use. We hope that the package will further the use of neighborhood competition models to better understand what structures plant competition.

@@

AYK and DNA conceived the ideas and coded a draft of the package. AYK wrote an initial manuscript draft. SPC rewrote much of the package's code to align with R and "tidy" best practices [@wickham_welcome_2019]. All authors contributed to subsequent drafts and gave final approval for manuscript.

### 2021-03-06 paper/forestecology.tex
first batch of edits

\bigskip
\begin{abstract}
1. Many models for the growth of trees that incorporate interspecies
competition are based on a neighborhood effect assumption whereby all
trees within a fixed distance of a focal tree are considered
competitors. Methods and tools are needed to quantify this competitive
effect and assess the quality of all resulting models.\\
2. We present the \texttt{forestecology} package providing methods for
both 1) evaluating the effect of competitor species identity using
permutation tests and 2) evaluating model performance using spatial
cross-validation. Following \citet{allen_permutation_2020}, we implement
a Bayesian linear regression competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We demonstrate that both
1) competitor species identity matters and 2) that not spatially
cross-validating leads to error estimates that are overly optimistic.\\
4. The package features 1) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, 2)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and 3) an S3 object-oriented implementation
of the Bayesian linear regression model. These three facts allow for
clear articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

\noindent%
{\it Keywords:} forest ecology, competition, tree growth, R, tidyverse, sf,
cross-validation, spatial statistics
\vfill

@@

BERT: Dave could you write a few sentences (and add references where
appropriate) starting with the big picture of modeling tree growth,
zooming in to neighhorhood models, and then connecting to the
``\citet{allen_permutation_2020} considers the following model'' I write
below?

Repeat-censused forest plots offer excellent data to test neighborhood
models of the effect of competition on the growth of trees
\citet{canham_neighborhood_2006} \citet{uriarte_spatially_2004}.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{growth_scbi <-}
\StringTok{  }\KeywordTok{compute_growth}\NormalTok{(}
    \DataTypeTok{census_1 =}\NormalTok{ census_}\DecValTok{2013}\NormalTok{_scbi,}
    \DataTypeTok{census_2 =}\NormalTok{ census_}\DecValTok{2018}\NormalTok{_scbi }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(}\OperatorTok{!}\KeywordTok{str_detect}\NormalTok{(codes, }\StringTok{"R"}\NormalTok{)),}
    \DataTypeTok{id =} \StringTok{"stemID"}
\NormalTok{  )}
\NormalTok{growth_scbi}
\CommentTok{## Simple feature collection with 7954 features and 8 fields}
\CommentTok{## geometry type:  POINT}
\CommentTok{## dimension:      XY}
\CommentTok{## bbox:           xmin: 0.2 ymin: 300 xmax: 300 ymax: 600}
\CommentTok{## CRS:            NA}
\CommentTok{## # A tibble: 7,954 x 9}
\CommentTok{##   stemID sp     dbh1 codes1 status  dbh2 codes2 growth}
\CommentTok{##    <dbl> <fct> <dbl> <chr>  <chr>  <dbl> <chr>   <dbl>}
\CommentTok{## 1      4 nysy  13.6  M      A       14.2 M       0.103}
\CommentTok{## 2      5 havi   8.8  M      A        9.6 M;P     0.150}
\CommentTok{## 3      6 havi   3.25 NULL   A        4   M       0.140}
\CommentTok{## 4     77 qual  65.2  M      A       66   M       0.141}
\CommentTok{## 5     79 tiam  47.7  M      A       46.8 M      -0.161}
\CommentTok{## # ... with 7,949 more rows, and 1 more variable: geometry <POINT>}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{focal_vs_comp_scbi}
\CommentTok{## # A tibble: 6,296 x 8}
\CommentTok{##   focal_ID focal_sp   dbh foldID   geometry growth comp             }
\CommentTok{##      <dbl> <fct>    <dbl> <fct>     <POINT>  <dbl> <list>           }
\CommentTok{## 1        4 nysy     13.6  1      (14.2 428)  0.103 <tibble [20 x 4]>}
\CommentTok{## 2        5 havi      8.8  1       (9.4 436)  0.150 <tibble [32 x 4]>}
\CommentTok{## 3       79 tiam     47.7  1        (40 381) -0.161 <tibble [20 x 4]>}
\CommentTok{## 4       80 caca      5.15 1      (38.7 422)  0.253 <tibble [12 x 4]>}
\CommentTok{## 5       96 libe      2.3  1        (60 310)  0.262 <tibble [14 x 4]>}
\CommentTok{## # ... with 6,291 more rows, and 1 more variable: growth_hat <dbl>}
\end{Highlighting}
\end{Shaded}

@@

Figure \ref{fig:scbi-simulation}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

The \texttt{forestecology} package provides

@@

AYK and DNA conceived the ideas and coded most of the package in the
early design stages. AYK led the writing of the manuscript. SPC
refactored much of the package's code in the later design stages to
align with R and ``tidy'' programming best practices
\citep{wickham_welcome_2019}. All authors contributed critically to the
drafts and gave final approval for publication.

>>>>

\bigskip
\begin{abstract}
1. Neighborhood competition models are powerful tools to measure the
effect of interspecific competition. Statistical methods to ease the
application of these models are currently lacking.\\
2. We present the \texttt{forestecology} package providing methods to i)
specify neighborhood competition models, ii) evalulate the effect of
competitor species identity using permutation tests, and iii) measure
model performance using spatial cross-validation. Following
\citet{allen_permutation_2020}, we implement a Bayesian linear
regression neighborhood competition model.\\
3. We demonstrate the package's functionality using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO global network of reseach sites. Given ForestGEO's
data collection protocols and data formatting standards, the package was
designed with cross-site compatibility in mind. We demonstrate that both
i) competitor species identity matters and ii) that not spatially
cross-validating leads to error estimates that are overly optimistic.\\
4. The package features i) \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence, ii)
functions with standardized inputs/outputs of simple features
\texttt{sf} package class, and iii) an S3 object-oriented implementation
of the Bayesian linear regression model. These three facts allow for
clear articulation of all the steps in the sequence of analysis and easy
wrangling and visualization of the geospatial forestry data.
Furthermore, while the package only has Bayesian linear regression
implemented, the package was designed with extensibility to other
methods in mind.
\end{abstract}

\noindent%
{\it Keywords:} forest ecology, interspecific competition, neigbhorhood competition,
tree growth, R, ForestGEO, cross-validation, spatial statistics,
\vfill

@@

Repeat-censused forest plots offer excellent data to test neighborhood
models of the effect of competition on the growth of trees
(\citet{canham_neighborhood_2004}). Neighborhood models of competition
have been used to: test whether the species identity of a competitor
matters (\citet{uriarte_spatially_2004}); measure species-specific
competition coefficients (\citet{das_effect_2012}
\citet{tatsumi_estimating_2016}); test competing models to see what
structures competitive interactions, e.g.~traits or phylogeny
(\citet{allen_permutation_2020} \citet{uriarte_trait_2010}); and inform
selective logging practices (\citet{canham_neighborhood_2006}). Although
these are well-described methods, no methods are currently available for
their easy application. Here we address this in an R package. We largely
follow the methods presented in \citet{allen_permutation_2020}. The
package is written to model stem radial growth between two censuses
based on neighborhood competition. It assumes a single large plot using
the ForestGEO plot data structure as a guide
(\citet{andersonteixeira_ctfs-forestgeo_2015}).

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{growth_scbi <-}
\StringTok{  }\KeywordTok{compute_growth}\NormalTok{(}
    \DataTypeTok{census_1 =}\NormalTok{ census_}\DecValTok{2013}\NormalTok{_scbi,}
    \DataTypeTok{census_2 =}\NormalTok{ census_}\DecValTok{2018}\NormalTok{_scbi }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(}\OperatorTok{!}\KeywordTok{str_detect}\NormalTok{(codes, }\StringTok{"R"}\NormalTok{)),}
    \DataTypeTok{id =} \StringTok{"stemID"}
\NormalTok{  )}
\NormalTok{growth_scbi}
\CommentTok{## Simple feature collection with 7954 features and 8 fields}
\CommentTok{## geometry type:  POINT}
\CommentTok{## dimension:      XY}
\CommentTok{## bbox:           xmin: 0.2 ymin: 300 xmax: 300 ymax: 600}
\CommentTok{## CRS:            NA}
\CommentTok{## # A tibble: 7,954 x 9}
\CommentTok{##   stemID sp     dbh1 codes1 status  dbh2 codes2 growth   geometry}
\CommentTok{##    <dbl> <fct> <dbl> <chr>  <chr>  <dbl> <chr>   <dbl>    <POINT>}
\CommentTok{## 1      4 nysy  13.6  M      A       14.2 M       0.103 (14.2 428)}
\CommentTok{## 2      5 havi   8.8  M      A        9.6 M;P     0.150  (9.4 436)}
\CommentTok{## 3      6 havi   3.25 NULL   A        4   M       0.140  (1.3 434)}
\CommentTok{## 4     77 qual  65.2  M      A       66   M       0.141 (34.7 307)}
\CommentTok{## 5     79 tiam  47.7  M      A       46.8 M      -0.161   (40 381)}
\CommentTok{## # ... with 7,949 more rows}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{focal_vs_comp_scbi}
\CommentTok{## # A tibble: 6,296 x 8}
\CommentTok{##   focal_ID focal_sp   dbh foldID                  geometry growth comp }
\CommentTok{##      <dbl> <fct>    <dbl> <fct>                    <POINT>  <dbl> <lis>}
\CommentTok{## 1        4 nysy     13.6  1                     (14.2 428)  0.103 <tib~}
\CommentTok{## 2        5 havi      8.8  1                      (9.4 436)  0.150 <tib~}
\CommentTok{## 3       79 tiam     47.7  1                       (40 381) -0.161 <tib~}
\CommentTok{## 4       80 caca      5.15 1                     (38.7 422)  0.253 <tib~}
\CommentTok{## 5       96 libe      2.3  1                       (60 310)  0.262 <tib~}
\CommentTok{## # ... with 6,291 more rows, and 1 more variable: growth_hat <dbl>}
\end{Highlighting}
\end{Shaded}

@@

Figure \ref{fig:scbi-simulation} highlights the importance of spatial
cross validating when interpreting results from neighborhood compeition
models. Without cross-validation the competitor idenity did matter, the
non-permuted competitor species had a much lower RMSE than the permuted
one. But once we include the spatial cross validation this improvement
disaprees. So in this 9 ha subplot of the SCBI plot competitive
interactions do not depend on the identity of the competitor, which is
the opposite of what has been observed in other places
(\citet{allen_permutation_2020} \citet{uriarte_spatially_2004}).

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The \texttt{forestecology} package provides an accessible way to fit and
test models of neighborhood competition. Currently it is written to work
with data from ForestGEO plots. But it could easily be modified to work
on any single large, mapped forest plot in which at least two
measurements of each individual have been taken. With a little bit of
work the package could also be applied to forest inventory data in which
several small plots are spatially separated, e.g., USFS Forest
Inventory. In a future version of the package we also hope to make it
possible to model plant mortality in addition to plant growth. The
package follows the guidelines for \texttt{tidy} data, \texttt{sf}
spatial structure., and S3 open-oriented model structure. These make for
ease of use. We hope that the package will further the use of
neighborhood competition models to better understand what structures
plant competition.

@@

AYK and DNA conceived the ideas and coded a draft of the package. AYK
wrote an initial manuscript draft. SPC rewrote much of the package's
code to align with R and ``tidy'' best practices
\citep{wickham_welcome_2019}. All authors contributed to subsequent
drafts and gave final approval for manuscript.
